### 2010-03-11 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: Code and interpretation face to face, for all Desc and IDesc definitions.
Desc@ICFP: Second section on Desc almost done. It is just too late, or too early to do initial algebras.
Desc@ICFP: close section Levitation.

\section{Levitating the universe of descriptions}

@@

\begin{structure}
<- Recall typing rules of 1st section
    -> Make clear they were just promises
    -> Can be implemented now
        <- Simply List UId
\end{structure}

\[\stk{
\EnumU : \Set \\
\EnumU \mapsto \Mu{(\List~\UId)}
}\]

\begin{structure}
<- Consequences
    -> Type theory doesn't need to be extended with EnumU, NilE, and ConsE
        <- EnumU == Mu EnumUD
        <- NilE, ConsE are just tags
    -> Do not need a specific \spi eliminator
        <- \spi is an instance of the generic eliminator
            <- Code?
    -> Anything else remains the same (switch, EnumT, 0, 1+)
\end{structure}

\begin{structure}
<- Summary of the operation
    <- The content of the type theory is exactly the same
        <- before == after
    /> type naming scheme condenses
        <- Replace named constructors by codes in the universe of data-types
    -> Our next step is a similar move (in essence)
        /> Condenses the entire naming scheme of data-types
\end{structure}

@@

\begin{structure}
<- Realizing our promises
    <- We are going to implement Desc
    /> Desc is itself a data-type
        <- Same situation as EnumU
            <- We want to benefit from generic operations
        -> It ought to be encoded in itself
\end{structure}

\subsubsection{First try}

\begin{structure}
<- A partial implementation
    <- '1 and 'indx are easy
    <- 'sigma is partially doable
        /> lack the ability to do an higher-order inductive call
    -> Show partial code [figure]
\end{structure}

\[\stk{
\DescD : \Desc \\
\begin{array}{@{}ll}
\DescD \mapsto \DSigma{}{} & (\EnumT [ \DUnit, \DSigma{}{}, \DIndx{} ]) \\
                           & \left[\begin{array}{l}
                                   \DUnit                                \\
                                   \DSigma{\Set}{(\LAM{\V{X}} ???)} \\
                                   \DIndx{\DUnit}                                  \\
                                   \end{array}
                             \right]
\end{array}
}\]

\subsubsection{Second try}

\begin{structure}
<- Extending the universe of description
    -> With higher-order induction
    <- Intuition: index elements in X by H, and go on reading
        -> indx is isomorph to hindx for H = 1
    /> Keep indx
        <- First order!
        -> Extensionally equal to hindx 1
        /> Practically, definitional equality on Sigma/Pi cannot cope with it
    -> Show DescD code
\end{structure}

\[
\stk{
\data \Desc : \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \ldots          & \:\:\ldots \\
    \DHindx         & \PI{H}{Set} \Desc \To \Desc
\end{array}
}
\]

\[\stk{
\descop{\_\:}{} : \Desc \To \Set \To \Set \\
\begin{array}{@{}l@{\:=\:\:}ll}
\ldots                        &  \ldots \\
\descop{\DHindx{H}{D}}{X}     &  \TIMES{(H \To X)}{\descop{D}{X}}
\end{array}
}\]

\[\stk{
\DescD : \Desc \\
\begin{array}{@{}ll}
\DescD \mapsto \DSigma{}{} & (\EnumT [ \DUnit, \DSigma{}{}, \DIndx{}, \DHindx{}{} ]) \\
                           & \left[\begin{array}{l}
                                   \DUnit                                \\
                                   \DSigma{\Set}{(\LAM{\V{X}} \DHindx{X}{\DUnit})} \\
                                   \DIndx{\DUnit}                                  \\
                                   \DSigma{\Set}{(\LAM{\_} \DIndx{\DUnit})}
                                   \end{array}
                             \right]
\end{array}
}
\]

@@

\begin{structure}
<- Subtlety: translation of [ ... ]
    -> Let us do it manually
        -> Code with problem for the motive of switch
\end{structure}

\begin{structure}
<- The magician trick
    <- Our problem is to give a motive for switch
        /> We perfectly know what it ought to be: \_ -> DescD
    -> Solution: extend the type theory with a special purpose switchD
        -> Only extension required to the type theory!
        -> Hidden away to the user by the syntactic sugar
            -> Sufficient to ensure unavailability as a raw operator
            <- Another instance of type propagation
\end{structure}

\begin{structure}
<- Generic programming now!
    <- Desc is just data
        -> Can be manipulated
    <- Free induction scheme on Desc
        -> Ability to inspect data-types
        -> Ability to program on data-types
\end{structure}

@@

\begin{structure}
<- Adding hindx have introduced some duplication
    <- indx == hindx 1
    -> We can factor out commonalities 
        /> Obtain an equivalent presentation
        /> Still embeddable (refer to the Agda model)
\end{structure}

\begin{structure}
<- Give new presentation [figure]
    <- hindx have introduced the notion of function space: 'Pi
    <- hindx and indx are both composed by a binary product and a left open term: 'x and 'id 
    -> Straightforward translation to the new system [equation]
\end{structure}

\[
\stk{
\data \Desc : \Set \where \\
\;\;\begin{array}{@{}l@{\::\:}l@{\quad}l}
    \DId            & \Desc                                   \\
    \DUnit          & \Desc                                   \\
    \DProd{}{}      & \PI{\V{D}, \V{D'}}{\Desc} \Desc         \\
    \DSigma         & \PI{\V{S}}{\Set} \PIS{S \To \Desc} \Desc \\
    \DPi            & \PI{\V{S}}{\Set} \PIS{S \To \Desc} \Desc 
\end{array}
}
\]

\[\stk{
\descop{\_\:}{} : \Desc \To \Set \To \Set \\
\begin{array}{@{}l@{\:=\:\:}ll}
\descop{\DId}{X}          &  X                                           \\
\descop{\DUnit}{X}        &  \Unit                                       \\
\descop{\DProd{D}{D'}}{X} &  \TIMES{\descop{D}{X}}{\descop{D'}{X}}       \\
\descop{\DSigma{S}{D}}{X} &  \SIGMA{\V{s}}{S} \descop{D\: s}{X}                \\
\descop{\DPi{S}{D}}{X}    &  \PI{\V{s}}{S} \descop{D\: s}{X}            
\end{array}
}\]

@@

@@

\begin{structure}
<- Making cata
    <- Present the type signature
    <- Starts with a call to generic induction
        <- induction on Desc!
        /> Show types at hand
        -> Explain how to use inductive hypothesis
    <- Implement the 'replace' function
    -> Dependent-typeless catamorphism 
\end{structure}

This \(\F{induction}\) operator is the natural dependent elimination
principle, but we might also benefit from the traditional \emph{catamorphism}
or `fold operator' which accompanies a (weakly) initial algebra. We should
like to have
\[\stk{
\F{cata} : \PITEL{D}{\Desc}
           \PI{T}{\Set}
           (\descop{D}{T}\To T) \To 
           \Mu{D} \To T \\
\F{cata}\: D\: T\: f \mapsto
  \F{induction}\: D\: (\LAM{\_}T)\: (\LAM{xs\:ts} f\: ?)
}\]
but what should \(?\) be? We have \(xs:\descop{D}{\Mu{D}}\)
and \(ts:\All{D}{(\Mu{D})}{(\LAM{\_}T)}{xs}\), so surely we can construct
an element of \(\descop{D}{T}\) by replacing each recursive component from
\(xs\) with its counterpart from \(ts\).

@@

\begin{structure}
<- Deriving generic functions
    <- Taking a Desc and computing a function
        <- Desc comes equipped with an induction principle
        -> Ability to compute more functions from it
            -> More generic functions
    <- Inspecting data-types
        <- All described by a Desc code
        -> Ability to explore the code
            <- Desc equipped with an induction principle
            -> Build new objects based on that structure
\end{structure}

@@

\begin{structure}
!!! EARLIER !!!
<- Tagged description
    <- Form TDesc = List (UId x Desc) [equation]
    <- Follow usual sums-of-product presentation of data-type
        <- Finite set of constructors
        <- Then whatever you want
    -> Any Desc data-type can be turned into this form
        -> No loss of expressive power
        /> Garantee a ``constructor form''
\end{structure}

A \emph{tagged} description is given by an inhabitant of
\[
 \F{TagDesc} \mapsto \List{\TIMES{\UId}{\Desc}}
\]

Datatypes specified in the conventional `sum of products' style naturally give rise to tagged descriptions. Of course, every description can be dorced into this style with a singleton choice of tag.

\begin{structure}
<- A generic program: the free monad construction
    <- Recall free monad construction in Haskell
        -> Based on a functor F
    <- Note that the free monad construction is itself defined by a functor
        -> Extract it
    <- Encode it in the Desc world [equation]
        <- F is the Desc we start with
        <- The free monad functor is what we have just defined
        <- [\_]* : Desc -> Set -> Desc
           [\_]* D X = 'cons ['var ('sigma X (\_ -> '1))] D
        -> Mu does the fix-point
\end{structure}

We may then implement the \emph{free monad} construction as a
transformation on \(\F{TagDesc}\).
\[\stk{
\FreeMonad{\_} : \F{TagDesc} \To \Set \To \F{TagDesc} \\
\FreeMonad{\pair{E}{D}{}}\:X \mapsto
\pair{\ListCons{\:\DVar{}}{E}}{\pair{\DConst{X}}{D}{}}{}
}\]
Of course, we must equip the resulting datatypes with operations delivering a monadic interface. As usual, \(\LAM{\x}\DVar{\x}\) performs the r\^ole of `return', embedding variables into terms. We shall also need a generic \emph{substitution} operator.

\begin{structure}
<- A generic program: monadic substitution [equation]
    <- subst : \forall T X Y. mu ([T]* X) -> (X -> mu ([T]* Y)) -> mu ([T]* Y)
        -> Using Fold
    -> Consequences
        <- We have free monad data-type
            <- Term + variables
        <- We have monad operations
            <- Return / var
            <- Substitution / bind
\end{structure}

\begin{structure}
<- Deriving new data-structure and functions on them
    <- Computing the Free Monad of a data-type
        <- Derive new data-structure from previous one
            <- It is just code
        /> New data-structure comes with some equipment
    <- Computing new functions on computed data-types
        <- If data comes with structure, we ought to be able to capture it
            <- Induction on Desc
            -> Ability to compute over data
\end{structure}

>>>>

\section{Levitating the universe of descriptions}
\label{sec:desc-levitate}

In this section, we will fulfill several promises. One promise we made
was to effectively implement the code of the $\Desc$ universe. Another
promise we tacitly made is to implement the type former of finite
sets. Despite being a perilous pedagogical exercise for the authors,
we hope to convey to the reader the dizzy feeling of levitation,
without the falling.

@@

\begin{wstructure}
<- Recall typing rules of 1st section
    -> Make clear they were just promises
    -> Can be implemented now
        <- Simply List UId
\end{wstructure}

In Section~\ref{sec:finite-sets}, we have promised an implementation
of finite sets. It consists in the objects specified in
Figure~\ref{fig:typing-finite-set}. In particular, we will now
implement the $\EnumU$ type former and its constructors. Let us recall
their specification:

\[\stkc{
%% EnumU
\Rule{\Gamma \vdash \Valid}
     {\Gamma \vdash \Type{\EnumU}} 
\\
%% NilE
\Rule{\Gamma \vdash \Valid}
     {\Gamma \vdash \Bhab{\NilE}{\EnumU}} 
\qquad
%% ConsE
\Rule{\Gamma \vdash \Bhab{t}{\UId} \quad
      \Gamma \vdash \Bhab{e}{\EnumU}}
     {\Gamma \vdash \Bhab{\ConsE{t}{e}}{\EnumU}}
}\]

The names $\NilE$ and $\ConsE{}{}$ are absolutely not a
coincidence. $\EnumU$ is indeed simply isomorph to a list, with
$\UId$s as parameter. However, in Section~\ref{sec:desc-examples}, we
have shown how to build lists in the universe of
description. Therefore, the actual implementation of $\EnumU$ sums up
to the following, trivial definition:

\[\stk{
\EnumU : \Set \\
\EnumU \mapsto \Mu{(\ListD~\UId)}
}\]

\begin{wstructure}
<- Consequences
    -> Type theory doesn't need to be extended with EnumU, NilE, and ConsE
        <- EnumU == Mu EnumUD
        <- NilE, ConsE are just tags
    -> Do not need a specific \spi eliminator
        <- \spi is an instance of the generic eliminator
            <- Code?
    -> Anything else remains the same (switch, EnumT, 0, 1+)
\end{wstructure}

Let us examine the consequences of our act. First of all, we discover
that the type theory does not actually need to be extended with the
type former $\EnumU$, the constructors $\NilE$ and $\ConsE$. We have
just defined $\EnumU$ above, by fix-point over the signature functor
$\ListD$. $\NilE$ and $\ConsE$ are simply the $\ListNil$ and
$\ListCons$ constructors of lists.

Another interesting consequence concerns the $\spi{}{}$ operator: it
can actually be implemented with the generic $\F{induction}$
principle. From an implementation perspective, this means that we
remove an operator, hard-coded in the host language. This operator is
simply implemented in the dependently-typed target language, as any
other function.

Apart from these changes, we are left with implementing the other
components of finite sets. This consists in $\EnumT$, $\Ze$, $\Su$, as
well as the $\F{switch}$ operator. Note that the actual implementation
of $\EnumU$ does not influence our implementation of $\EnumT$: be they
hard-coded or codes in the $\Desc$ universe, the $\EnumU$ objects just
behave similarly. We have witnessed this effect when carrying this
operation in Epigram, moving from an hard-coded presentation to a
self-hosted one. Absolutely no changes to the $\EnumT$ world was
required.

\begin{wstructure}
<- Summary of the operation
    <- The content of the type theory is exactly the same
        <- before == after
    /> type naming scheme condenses
        <- Replace named constructors by codes in the universe of data-types
    -> Our next step is a similar move (in essence)
        /> Condenses the entire naming scheme of data-types
\end{wstructure}

In this section, we have replaced a low-level presentation of finite
sets by a self-hosted one, expressed in the universe of
descriptions. However, formally, the content of the type theory
remains unchanged: objects which were present before the modification
are still there. Conversely, we have not introduced any spurious
objects in the type theory.

If not on the content, this modification had an effect on the
names. The type naming scheme of the type theory has condensed: named
type formes ($\EnumU$) and constructors ($\NilE$ and $\ConsE$) are now
replaced by codes and their fix-point in the universe of
descriptions. Our next step is similar, in essence: we are going to
condense the entire naming scheme of data-types \emph{in itself}.

@@

\begin{wstructure}
<- Realizing our promises
    <- We are going to implement Desc
    /> Desc is itself a data-type
        <- Same situation as EnumU
            <- We want to benefit from generic operations
        -> It ought to be encoded in itself
\end{wstructure}

We shall now fulfill our promises. We are going to implement the codes
of the universe of descriptions. Interestingly, in and by itself, a
code is nothing but a data-type. We are in the same situation than
with $\EnumU$: we ought to be able to describe the codes of $\Desc$ in
$\Desc$ itself. Hence, this code would be a first-class citizen, born
with the standard, generic equipment of data-types.

\subsubsection{First attempt}

\begin{wstructure}
<- A partial implementation
    <- '1 and 'indx are easy
    <- 'sigma is partially doable
        /> lack the ability to do an higher-order inductive call
    -> Show partial code [figure]
\end{wstructure}

Our first tentative is the following:

\[\stk{
\DescD : \Desc \\
\begin{array}{@{}ll}
\DescD \mapsto \DSigma{}{} & (\EnumT [ \DUnit, \DSigma{}{}, \DIndx{} ])  \\
                           & \left[\begin{array}{l}
                                   \DUnit                                \\
                                   \DSigma{\Set}{(\LAM{\V{S}} ???)}      \\
                                   \DIndx{\DUnit}                        \\
                                   \end{array}
                             \right]
\end{array}
}\]

Let us explain how we have proceeded and the obstacle we face. Much as
the data-types we have seen so far, $\DescD$ first consists in chosing
a constructor. Constructors correspond to the code's cases: $\DUnit$,
$\DSigma$, and $\DIndx$. The reader will notice that the tagged
notation we have used so far for $\Desc$ constructors now fully make
sense: these were actually the tags we are defining.

The case of $\DUnit$ is trivial: we immediately reach the end of the
data-type. $\DUnit$ has no payload. The case of $\DIndx{}$ comes
naturally: it only takes a recursive argument -- a description. The
case of $\DSigma$ is problematic. Recall the definition of $\DSigma$:

\[    \DSigma{}{} : \PI{\V{S}}{\Set} \PIS{S \To \Desc} \Desc      \]

So, we first pack an element $S$ of $\Set$. Then, we would like to
express a notion of recursive argument \emph{indexed} by $S$. Because
our presentation is entirely first-order so far, we are not able to
express this notion. We shall remedy to this situation.

\subsubsection{Second attempt}

\begin{wstructure}
<- Extending the universe of description
    -> With higher-order induction
    <- Intuition: index elements in X by H, and go on reading
        -> indx is isomorph to hindx for H = 1
    /> Keep indx
        <- First order!
        -> Extensionally equal to hindx 1
        /> Practically, definitional equality on Sigma/Pi cannot cope with it
    -> Show DescD code
\end{wstructure}

In order to capture a notion of higher-order induction, we need to
extend our universe of descriptions. This consists in adding a code
$\DHindx$ that takes an indexing set $H$. The code and its
interpretation are presented in
Figure~\ref{fig:hindx_desc}. Intuitively, $\DHindx$ use the elements
of $H$ to index as many recursive arguments.

\begin{figure*}

\[
\begin{array}{ll}
\stk{
\data \Desc : \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \ldots          & \:\:\ldots \\
    \DHindx         & \PI{H}{Set} \Desc \To \Desc
\end{array}
}
&
\stk{
\descop{\_\:}{} : \Desc \To \Set \To \Set \\
\begin{array}{@{}l@{\:=\:\:}ll}
\ldots                        &  \ldots \\
\descop{\DHindx{H}{D}}{X}     &  \TIMES{(H \To X)}{\descop{D}{X}}
\end{array}
}
\end{array}
\]

\caption{Higher-order universe of descriptions}
\label{fig:hindx_desc}

\end{figure*}

Note that $\DIndx$ is isomorph to $\DHindx{\Unit}$. However, we
tolerate this duplication and do not replace $\DIndx$ by the more
general $\DHindx$. Indeed, unlike its counterpart, $DIndx$ is
first-order. Therefore, while both codes are \emph{extensionally} the
same, in most practical implementation, they would be dealt with
rather differently. While we can definitional equality to cope with
first order objects, the functional presentation introduced by
$\DHindx$ is unlikely to be amenable to a purely definitional
treatment.

Equipped with $\DHindx$, we can describe the extended code of our
universe of data-types, as shown in Figure~\ref{fig:desc-levitate}.
The $\DUnit$ and $\DIndx$ cases remains unchanged, as expected. We
achieve the description of the $\DSigma$ case, by a simple appeal to
the higher-order induction on $S$. The $\DHindx$ case consists in
packing the $H$ in $\Set$ with a recursive arguement.

In a first glance, we have achieved our goal. We have described the
codes of the universe of description. Taking the fix-point of this
object gives us $\Desc$, up to isomorphism. We have implemented our
code of data-types as a levitating object inside itself. However, this
levitation operation, just as any magic trick, relies on an invisible
cable. Let us reveal it, hence finishing our implementation.

\begin{figure}

\[\stk{
\DescD : \Desc \\
\begin{array}{@{}ll}
\DescD \mapsto \DSigma{}{} & (\EnumT [ \DUnit, \DSigma{}{}, \DIndx{}, \DHindx{}{} ]) \\
                           & \left[\begin{array}{l}
                                   \DUnit                                            \\
                                   \DSigma{\Set}{(\LAM{\V{S}} \DHindx{S}{\DUnit})}   \\
                                   \DIndx{\DUnit}                                    \\
                                   \DSigma{\Set}{(\LAM{\V{H}} \DIndx{\DUnit})}
                                   \end{array}
                             \right]
\end{array}
}\]

\caption{Levitating description of $\Desc$}
\label{fig:desc-levitate}

\end{figure}

@@

\begin{wstructure}
<- Subtlety: translation of [ ... ]
    -> Let us do it manually
        -> Code with problem for the motive of switch
\end{wstructure}

Our reader might be slightly confused to learn that the trick is
visible in the definition of $\DescD$
(Fig.\ref{fig:desc-levitate}). Or rather, it is made invisible by a
careful usage of type propagation. Indeed, let us elaborate this term
down to the low-level type theory. The $\EnumT [ \ldots ]$ construct
elaborates to a finite set $e$ in $\EnumU$, inhabited by the
codes. Then, we can type-check the case definition, between square
brackets $[ \ldots ]$. This term $f$ is pushed into the type
$\EnumT{e} \To \Desc$. This corresponds to a finite function set we
have implemented in Figure~\ref{fig:type-checking}.

What happens if we unfold the definition? We ought to build the following term:

\[
\PLAM{x}{(\EnumT{e})} \switch{e}{\PLAM{\_} \Desc}{\pi^f}{x}
\]

But this is rather problematic! We are still in the process of
constructing $\DescD$, and the motive of $\F{switch}$ is abruptely
begging for an $\eta$-expanded $\Desc$. Despite our willingness, we
will never be able to materialize such motive.

\begin{wstructure}
<- The magician trick
    <- Our problem is to give a motive for switch
        /> We perfectly know what it ought to be: \_ -> DescD
    -> Solution: extend the type theory with a special purpose switchD
        -> Only extension required to the type theory!
        -> Hidden away to the user by the syntactic sugar
            -> Sufficient to ensure unavailability as a raw operator
            <- Another instance of type propagation
\end{wstructure}

If we cannot make it happen, we simply do not make it happen. We
perfectly know what the motive ought to be. So, straightforwardly, we
extend the type theory with two special-purpose operators:

\[
\begin{array}{@{}ll}
%% piD
\green{\pi_D} : & \PITEL{\V{e}}{\EnumU} \To \Set \\
%% switchD
\F{switchD} : & \PITEL{\V{e}}{\EnumU}               
                \PITEL{\V{b}}{\spiD{e}}
                \PITEL{\V{x}}{\EnumT{e}} \To \Desc
\end{array}
\]

The entire work of the magician stands here, in these two
extensions. One could be worried by the availability of such
operations in our type theory. Indeed, it would be inconvenient for
the user to have to decide between these two variants. In fact, we
give no choice to the user, hence no inconvenience. Just as the
$\F{switch}$ operator was sealed off by the type propagation
machinery, $\F{switchD}$ is not provided to the user. We simply extend
the type-checking rules (Fig.~\ref{fig:type-checking}) with the
following inference rule:

\[
\Rule{\Gamma \Vdash \propag{\push{t}{\spiD{e}}}
                           {t'}}
     {\Gamma \Vdash \begin{array}{@{}l} 
                        \propag{\push{t}{\EnumT{e} \To \Desc}}
                               {\\ \PLAM{x}{(\EnumT{e})} \switchD{e}{t'}{x}}
                    \end{array}}\;\mbox{t is $[]$ or $[a,b]$}
\]

\begin{wstructure}
<- Generic programming now!
    <- Desc is just data
        -> Can be manipulated
    <- Free induction scheme on Desc
        -> Ability to inspect data-types
        -> Ability to program on data-types
\end{wstructure}

%%! Maybe a bit of perspective would be good
%%    <- Everything stands up with the Mu
%%    <- Induction principle defined over Mu-things

This concludes our levitation work. Beyond the beauty of the exercise,
there is also several practical outcomes. First of all, this exercise
reveals that the $\Desc$ universe is just plain data. Just as any
piece of data, it can therefore be inspected and manipulated. However,
unlike any data, it is expressed in the $\Desc$ universe. As a
consequence, it is equipped, for free, with an induction
principle. So, our ability to inspect and program with $\Desc$ is not
restricted to the meta-language: we now have all the necessary
equipment in the target language to \emph{program} over data-types. In
this setting, generic programming is just about \emph{programming}.

@@

\begin{wstructure}
<- Adding hindx have introduced some duplication
    <- indx == hindx 1
    -> We can factor out commonalities 
        /> Obtain an equivalent presentation
        /> Still embeddable (refer to the Agda model)
\end{wstructure}

Before moving on some examples of generic programming, we have to
carry out some maintenance work. The extension of $\Desc$ with
$\DHindx$ have introduced some duplication of functionality. In order
to remove this clutter, we will factor out these commonalities. This
results in a new definition of the $\Desc$ universe. This
reorganization is strictly equivalent to the previous one
(Fig.~\ref{fig:hindx_desc}). Just as the previous version, it is also
self-descriptive. We refer the reader to our companion technical
report for details.

\begin{wstructure}
<- Give new presentation [figure]
    <- hindx have introduced the notion of function space: 'Pi
    <- hindx and indx are both composed by a binary product and a left open term: 'x and 'id 
    -> Straightforward translation to the new system [equation]
\end{wstructure}

The factorized universe of description is presented in
Figure~\ref{fig:factor-desc}. The modifications are justified as
follow. First, $\DHindx$ have introduced a notion of function space:
we capture this notion by the $\DPi$ constructor. Second, both
$\DHindx$ and $\DIndx$ are composed by a product and the functorial
argument: we capture these two notions with $\DProd{}{}$ and $\DId$,
respectively.

In this finer-grained presentation, we can define $\DIndx$ and
$\DHindx$ as follow:

@@

Consequently, the examples developped above can be straightforwardly
translated into this new presentation. In the following, we adopt this
last version as our de facto universe of inductive types. In
particular, we are going to demonstrate several instances of generic
programming.

%%! Shall we talk about the Type Theory being Desc Zero? or such story?

\begin{figure*}

\[
\begin{array}{ll}
\stk{
\data \Desc : \Set \where \\
\;\;\begin{array}{@{}l@{\::\:}l@{\quad}l}
    \DId            & \Desc                                   \\
    \DUnit          & \Desc                                   \\
    \DProd{}{}      & \PI{\V{D}, \V{D'}}{\Desc} \Desc         \\
    \DSigma         & \PI{\V{S}}{\Set} \PIS{S \To \Desc} \Desc \\
    \DPi            & \PI{\V{S}}{\Set} \PIS{S \To \Desc} \Desc 
\end{array}
}
&
\stk{
\descop{\_\:}{} : \Desc \To \Set \To \Set \\
\begin{array}{@{}l@{\:=\:\:}ll}
\descop{\DId}{X}          &  X                                           \\
\descop{\DUnit}{X}        &  \Unit                                       \\
\descop{\DProd{D}{D'}}{X} &  \TIMES{\descop{D}{X}}{\descop{D'}{X}}       \\
\descop{\DSigma{S}{D}}{X} &  \SIGMA{\V{s}}{S} \descop{D\: s}{X}                \\
\descop{\DPi{S}{D}}{X}    &  \PI{\V{s}}{S} \descop{D\: s}{X}            
\end{array}
}
\end{array}
\]

\caption{Factorized universe of descriptions}
\label{fig:factor-desc}

\end{figure*}

@@

\begin{wstructure}
<- Making cata
    <- Present the type signature
    <- Starts with a call to generic induction
        <- induction on Desc!
        /> Show types at hand
        -> Explain how to use inductive hypothesis
    <- Implement the 'replace' function
    -> Dependent-typeless catamorphism 
\end{wstructure}

In Section~\ref{sec:desc-fix-point}, we have argued for the
implementation of a dependently-typed $\F{induction}$ principle,
instead of the more traditional catamorphism. However, in some
circumstances, the full-power of a dependent elimination is not
necessary. In the following, we propose to derive the catamorphism
from the generic induction principle.

The catamorphism can be characterized as an induction on the
description $D$, with a non-dependent motive targetting $T$. Given a
node $xs$ and the induction hypotheses, the method ought to build an
element of $T$. Provided we know how to make an element of
$\descop{D}{T}$, this step will be performed by the algebra $f$. Let
us take a look at this jigsaw:

\newcommand{\cata}{\F{cata}}

\[\stk{
\cata : \PITEL{D}{\Desc}
           \PI{T}{\Set}
           (\descop{D}{T} \To T) \To 
           \Mu{D} \To T \\
\cata\: D\: T\: f \mapsto
  \F{induction}\: D\: (\LAM{\_}T)\: (\LAM{xs\:hs} f\: ???)
}\]

We are left with implementing the \(???\) program. Recall that we have
\(\Bhab{xs}{\descop{D}{\Mu{D}}}\) and
\(\Bhab{hs}{\All{D}{(\Mu{D})}{(\LAM{\_} T)}{xs}}\) at hand. Our goal
is to make an element of \(\descop{D}{T}\). Intuitively, $xs$ is of
the right shape, but its sub-elements are of the wrong type. On the
other hand, for each sub-element of $xs$, $hs$ gives us the
corresponding element in $T$. Hence the name ``induction
hypotheses''. Therefore, to construct an element of \(\descop{D}{T}\),
we replace the recursive components of \(xs\) by their counterpart in
\(hs\):

%%! Update to the factorized presentation

@@

Filling the \(???\) with \(\F{replace}\: D\: (\Mu{D})\: T\: xs\: hs\) closes the
problem. In the type theory, we have built a generic catamorphism. Any
data-type will now come equipped with this operation, for free.

\begin{wstructure}
<- Deriving generic functions
    <- Taking a Desc and computing a function
        <- Desc comes equipped with an induction principle
        -> Ability to compute more functions from it
            -> More generic functions
    <- Inspecting data-types
        <- All described by a Desc code
        -> Ability to explore the code
            <- Desc equipped with an induction principle
            -> Build new objects based on that structure
\end{wstructure}

With this example, we have shown how we can derive a generic
operation, the catamorphism, from a pre-existing generic operation,
the induction principle. This has been made possible by our ability to
manipulate descriptions as first-class objects: the catamorphism is,
basically, a function mapping a $\Desc$ to a data-type specific
operation. Having the ability to program on $\Desc$ is the key to
define such function. The induction principle of $\Desc$, inherited by
levitation, plays a central role here: without an induction principle,
it would be impossible to program on $\Desc$.

Moreover, the $\F{replace}$ function demonstrates the benefit of an
approach based on universes. The data-types living in the universe of
descriptions, we are able to \emph{inspect} them. As shown by
$\F{replace}$, it is easy to explore these structures, as well as
building new ones.

@@

\begin{wstructure}
<- A generic program: the free monad construction
    <- Recall free monad construction in Haskell
        -> Based on a functor F
    <- Note that the free monad construction is itself defined by a functor
        -> Extract it
\end{wstructure}

In this section, we will turn to a more ambitious generic operation on
data-type. Given a functor, represented as a tagged description, we
build the free monad over this functor.

\newcommand{\FMFreeMonad}{\D{FreeMonad}}
\newcommand{\FMFreeMonadF}{\D{FreeMonad$_{\CN{Free}}$}}
\newcommand{\FMVar}{\C{Var}}
\newcommand{\FMComposite}{\C{Composite}}

Let us recall the free monad construction. Given a functor $F$, the
free monad over $F$ is defined by the following data-type:

\[
\stk{
\data \FMFreeMonad : \PITEL{\V{F}}{\Set \To \Set} 
                     \PI{\V{X}}{\Set} 
                     \Set 
\where \\
\;\;\begin{array}{@{}l@{\::\:}l@{\quad}l}
    \FMVar           & X \To \FMFreeMonad\: F\: X                            \\
    \FMComposite     & F (\FMFreeMonad\: F\: X) \To \FMFreeMonad\: F\: X    
\end{array}
}
\]

%%! Recall monadic structure of this object

Being an inductive type, this $\FMFreeMonad$ data-type is itself
defined by a functor signature. It is given by:

\[
\FMFreeMonadF\: F\: X\: Z \mapsto X + F Z
\]

\begin{wstructure}
    <- Encode it in the Desc world [equation]
        <- F is the Desc we start with
        <- The free monad functor is what we have just defined
        <- [\_]* : Desc -> Set -> Desc
           [\_]* D X = 'cons ['var ('sigma X (\_ -> '1))] D
        -> Mu does the fix-point
\end{wstructure}

In our setting, the functor will be represented by a tagged
description. The free monad construction will take such tagged
description, a set $X$ of variables, and will compute the tagged
description of the corresponding free monad.

Implementing this function is surprisingly easy:

\[\stk{
\FreeMonad{\_} : \TagDesc \To \Set \To \TagDesc \\
\FreeMonad{\pair{E}{D}{}}\:X \mapsto
    \pair{\ListCons{\DVar{}}{E}}
         {\pair{\DSigma{X}{\Unit}}{D}{}}{}
}\]

We simply add a constructor, $\DVar{}$, and define its argument to be
a $\DSigma{X}{\Unit}$, that is an element of $X$. We keep $E$ and $D$
as they were, hence leaving the functor untouched. The interpretation
of this definition convince us that we have built the functor
signature of the free monad. The fix-point operation ties the knot and
gives us the full-blown free monad construction.

\begin{wstructure}
<- A generic program: monadic substitution [equation]
    <- subst : \forall T X Y. mu ([T]* X) -> (X -> mu ([T]* Y)) -> mu ([T]* Y)
        -> Using Fold
\end{wstructure}

Of course, we must equip the resulting datatypes with operations
delivering a monadic interface. As usual, \(\LAM{\x}\DVar{\x}\)
performs the r\^ole of ``return'', embedding variables into terms. The
``bind'' operation corresponds to \emph{substitution}. We will now
implement it, as a generic function.

Our implementation will appeal to the $\cata$ function developped
previously. So, let us write down the types, and fill as much
arguments to $\cata$ as possible:

\newcommand{\subst}{\F{subst}}
\newcommand{\apply}{\F{apply}}

%%! This is not quite Mu of a tagged description

\[\stk{
\begin{array}{@{}ll}
\subst : & \PITEL{\V{D}}{\TagDesc}
           \PI{\V{X}, \V{Y}}{\Set} \\
         & \Mu{(\FreeMonad{D}{X})} \To
           (\V{X} \To \Mu{(\FreeMonad{D}{Y})}) \To
           \Mu{(\FreeMonad{D}{Y})} 
\end{array} \\
\subst\: D\: X\: Y\: x\: \sigma \mapsto
  \cata\: (\FreeMonad{D}{X})\: (\Mu{(\FreeMonad{D}{Y})})\: ???\: x
}\]

Our task is therefore to implement the algebra of the
catamorphism. Intuitively, its role is to catch appearance of
$\DVar{x}$, to replace it by $\sigma x$. Otherwise, it should simply
appeal to the induction hypothesis. This corresponds to the following
definition:

%%! Triple check. That doesn't work, right?

\[\stk{
\begin{array}{@{}ll}
\apply : & \PITEL{\V{D}}{\TagDesc} 
           \PI{\V{X}, \V{Y}}{\Set} \\
         & (\V{X} \To \Mu{(\FreeMonad{D}{X})}) \To
           \descop{\FreeMonad{D}{X}}{\Mu{(\FreeMonad{D}{Y})}} \To
           \Mu{(\FreeMonad{D}{Y})}                            \\
\end{array} \\
\begin{array}{@{}l@{\:\:\mapsto\:}l}
\apply\: \DVar{}\: X\: Y\: \sigma\: x  & \sigma x     \\
\apply\: xs\:      X\: Y\: \sigma\: ys & ys
\end{array}
}\]

\begin{wstructure}
    -> Consequences
        <- We have free monad data-type
            <- Term + variables
        <- We have monad operations
            <- Return / var
            <- Substitution / bind
\end{wstructure}

Filling the $???$ subgoal with $\apply\: D\: X\: Y\: \sigma$ finishes
the implementation. To sum up, we have implemented the free monad
construction for an arbitrary tagged description. This gives our user
the ability, for any data-type, to extend it with a notion of
variable. Then, with have equipped this structure with the
corresponding monadic operation, ``bind'' and ``return''. Whereas the
``return'' is a trivial variable former, the ``bind'' operation
corresponds to the substitution of variables by terms. We have shown
how to implement it, for any free monad.

\begin{wstructure}
<- Deriving new data-structure and functions on them
    <- Computing the Free Monad of a data-type
        <- Derive new data-structure from previous one
            <- It is just code
        /> New data-structure comes with some equipment
    <- Computing new functions on computed data-types
        <- If data comes with structure, we ought to be able to capture it
            <- Induction on Desc
            -> Ability to compute over data
\end{wstructure}
